\Large The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events

### 2023-09-13 paper/main.tex
working on integrating new studies into paper [WIP]

\title{\Large The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events}
\author[1]{Xinming Xu}
\author[2]{Ziyan Zhu}
\author[1, $\star$]{Jeremy R. Manning}
\affil[1]{Dartmouth College, Hanover, NH, USA}
\affil[2]{Peking University, Beijing, China}
\affil[$\star$]{Address correspondence to jeremy.r.manning@dartmouth.edu}

\newcommand{\stimDescription}{S1} % table
\newcommand{\segments}{S1} % figure
\newcommand{\events}{S2} %
\newcommand{\references}{S3}

@@

\begin{abstract} {
  How much can we infer about the past and future, given our knowledge of the present?  Unlike temporally symmetric inferences about simple sequences, inferences about our own lives are asymmetric: we are better able to infer the past than the future, since we remember our past but not our future (i.e., the psychological arrow of time).  What happens when both the past and future are unobserved, as when we make inferences about \textit{other} people's lives?  We had participants view segments of a character-driven television drama.  They wrote out what would happen just before or after each just-watched segment.  Participants were better at inferring past (versus future) events.  This asymmetry was driven by participants’ reliance on characters’ conversational references in the narrative, which tended to favor the past.  Our work reveals a temporal asymmetry in how observations of other people’s behaviors can inform us about the past and future.

\textbf{Keywords: arrow of time, prediction, retrodiction, narrative, conversation}}

@@

One key difference between simple artificial sequences and more naturalistic (real or narrative) sequences is that naturalistic sequences often incorporate other people.  Despite the past and future being equally unknown to \textit{the observer} prior to the current moment, other people, and realistic characters in narratives, have their own psychological arrows of time. Specifically, they have memories of their own pasts. Other people's asymmetric knowledge about their \textit{own} pasts and futures might affect their behaviors (e.g., conversations). In turn, this might provide time-asymmetric clues that favor the past~\citep[e.g., other people might talk more about their own pasts than their futures; ][]{DemiEtal18}. If observers leverage these clues from other people's asymmetric knowledge, then observers should also be better at inferring the past (versus the future) of other people's lives.  Alternatively, if inferences about other people's lives are more like inferences about artificial statistical sequences~\citep[e.g., perhaps solely relying on statistical regularities like event schemas, scripts, or situation models][]{RadvCope06, ZwaaRadv98, BoweEtal79, RangRitc12}, then the accuracy of inferences about the past and the future of others' lives should be approximately equal.

We designed a naturalistic paradigm for exposing participants to scenarios where the past and future were equally unobserved.  We asked our participants to watch a series of movie segments drawn from a character-driven dramatic television show.  Across the conditions and trials in the experiment, participants made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment.  We used manual annotations and sentence-level natural language processing models to characterize participants' responses.  To foreshadow our results, we found that participants were overall better at retrodicting the past than predicting the future. This appeared to be driven by two main factors.  First, characters more often referred to past events than future (e.g., planned) events, and this influenced participants' responses. Second, associations and dependencies between temporally adjacent events enabled participants to form estimates about nearby events (e.g., to a just-watched scene or a past or future event referenced in an observed conversation).  Taken together, our work reveals a temporal asymmetry in how observations of other humans’ behaviors inform us about the past versus the future.

\section*{Results}
Participants in our study ($n = 36$) watched segments from two storylines, drawn from the CBS television show \textit{Why Women Kill}.  Each storyline comprised 11 segments (mean duration: 2.05~min; range: 0.97--3.87~min, Table \stimDescription).  We asked participants to use free-form (typed) text responses to retrodict what had happened prior to a just-watched segment, predict what would happen next, or recall what they had just watched (Fig.~\ref{fig:method}, \textit{Task design}). We referred to the to-be-retrodicted, to-be-predicted, or to-be-recalled segment as the \textit{target segment} for each response. We systematically varied whether participants watched the segments in forward or reverse chronological order, and how many segments they had seen prior to making a response (see \textit{Methods}).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{methods}
  \caption{\textbf{Task overview.} Participants watched segments of two storylines from the television series \textit{Why Women Kill}.  They made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment. Across four counterbalanced sequences, we systematically varied whether participants watched the segments in forward or reverse chronological order, whether (or not) responses were cued using the main characters in the target segment, and which other segments participants had watched prior to making a response.  For each segment, we collected several retrodiction, prediction, and/or recall responses across different experimental conditions.  Experiment time is denoted along the
  vertical axis, storyline segments are indicated along the horizontal axis, and the colors denote experimental tasks (conditions).}
  \label{fig:method}
\end{figure}

We asked participants to generate four types of responses after watching each video segment: uncued responses, character-cued responses, updated responses, and recalls (Fig.~\ref{fig:method}, \textit{Data overview}).  To generate \textit{uncued} responses, we asked participants to either retrodict (uncued retrodiction; \textit{u-R}) what happened shortly before or predict (uncued prediction; \textit{u-P}) what happened shortly after the just-watched segment.  To generate \textit{character-cued} responses, we asked participants to retrodict (character-cued retrodiction; \textit{c-R}) or predict (character-cued prediction; \textit{c-P}) what came before or after the just-watched segment, but we provided additional information to the participant about which character(s) would be present in the target (to-be-retrodicted or to-be-predicted) segment.  We hypothesized that character-cued responses should be more accurate than uncued responses, to the extent that participants incorporate the character information we provided to them into their retrodictions and predictions.  To generate updated responses, we asked participants to watch an additional segment that came just prior to or just after the target segment, and then to update their retrodiction (\textit{c-RP}) or prediction (\textit{c-PR}) about the target segment. Results on updated responses are not reported in this paper. Finally, we also asked participants to \textit{recall} what happened in the just-watched segment.  We labeled these responses according to which other segments participants had watched prior to the just-watched target.  Retrodiction-matched recall (\textit{re(R)}) responses were made during the retrodiction sequences (B1 and B2; Fig.~\ref{fig:method}), whereas prediction-matched recall (\textit{re(P)}) responses were made during the prediction sequences (A1 and A2; Fig.~\ref{fig:method}).  Whereas retrodiction and prediction responses reflect what participants \textit{estimate} they would remember after watching the (inferred) target segment, recall responses provide a benchmark for comparison by measuring what they \textit{actually} remember about the target segment.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.80\textwidth]{results1}
  \caption{\textbf{Retrodiction, prediction, and recall performance by experimental condition.} \textbf{A.  Methods schematic.}  For each retrodiction, prediction, and recall response, we calculated the hit rate for events in the target segment, the response precision (see \textit{Methods}), and the response convergence across participants (see \textit{Methods}).  \textbf{B. Target event hit rate.} Mean proportions of target events that were contained in participants' responses, for each response type, averaged across target segments.  \textbf{C. Response precision.}  Mean precisions of participants' responses, for each response type, averaged across target segments.  The horizontal lines denote the mean pairwise semantic similarities (see \textit{Methods}) across recall responses (re(R): orange; re(P): blue).  \textbf{D. Response convergence.}  Mean (across-participant) convergence of participants' responses, for each response type, averaged across target segments.  All panels: error bars denote bootstrapped 95\% confidence intervals.  Asterisks indicate significance in the (generalized) linear mixed models: * denotes $p < 0.05$ and ** denotes $p < 0.01$.}
  \label{fig:result1}
\end{figure}

We used two general approaches to assess the quality of participants' responses (see \textit{Methods}, Fig.~\ref{fig:result1}A).  One approach entailed manually annotating events in the video and counting the number of matched events in participants' responses.  We identified a total of 117 unique events reflected across the 22 video segments (range: 3--9 per segment; see \textit{Methods}, Table \stimDescription).  We assigned one ``point'' to each of these video events.  We also identified 23 additional events in participants' responses that were either summaries of several events or that were partial matches to the manually identified video events.  We assigned 0.5 point to each of these additional events.  This point system enabled us to compute the numbers and proportions (\textit{hit rates}) of correctly retrodicted, predicted, and recalled events contained in each response.  Our second approach entailed using a natural language processing model~\citep{CerEtal18} to embed annotations and responses in a 512-dimensional feature space.  This approach was designed to capture conceptual overlap between responses that were not necessarily tied to specific events.  To quantify this conceptual overlap, we computed the similarities between the embeddings of different sets of responses.  Following \cite{HeusEtal21}, we defined the \textit{precision} of each participants' retrodictions or predictions about a target segment as the median cosine similarities between the embeddings of (a) the participant's retrodiction or prediction response for the target segment and (b) each \textit{other} participant's recalls of the same segment.  In other words, precision is designed to measure the extent to which retrodictions and predictions captured the conceptual content that (other) participants remembered.  We also developed a related measure, which we call \textit{convergence}, to characterize response similarities across participants.  In particular, we defined convergence as the mean cosine similarity between the embeddings of a participant's responses to a target segment and all other participants' responses (of the same type) to the same segment.  We analyzed the data using generalized linear mixed models, with participant and stimulus (e.g., target segment) identities as crossed random effects (see \textit{Methods}).

First we sought to validate a main effect of response type (i.e., uncued responses, character-cued responses, and recalls), irrespective of the temporal direction (retrodiction versus prediction).  Across these three types of responses, participants have access to increasing amounts of information about the target segment.  Therefore, across these response types, we hypothesized that participants' responses should become both more accurate and more convergent across individuals.  Consistent with this hypothesis, participants' character-cued retrodictions and predictions were associated with higher target event hit rates than uncued retrodictions and predictions (odds ratio (OR): 2.65, $Z = 4.24$, $p < 0.001$, 95\% confidence interval (CI): 1.69 to 4.16; Fig.~\ref{fig:result1}B).  These character-cued responses were also more precise ($b = 0.13$, $t(18.1) = 9.43$, $p < 0.001$, CI: 0.10 to 0.16; Fig.~\ref{fig:result1}C) and convergent across individuals ($b = 0.11$, $t(18.6) = 6.21$, $p < 0.001$, CI: 0.07 to 0.15; Fig.~\ref{fig:result1}D).   Relative to character-cued responses, participants' recalls showed higher target event hit rates (OR = 21.83, $Z = 10.61$, $p < 0.001$, CI: 12.35 to 38.59) and were more convergence across individuals ($b = 0.20$, $t(19.4) = 9.10$, $p < 0.001$, CI: 0.16 to 0.25). These results are consistent with the common-sense notion that access to more information about a target segment yields better performance (i.e., higher hit rates, precision, and convergence across individuals).

Next we carried out a series of analyses specifically aimed at characterizing temporal direction effects--- i.e, the relative quality of retrodictions versus predictions across different types of responses.  We hoped that these analyses might provide insights into our central question about whether inferences about the past and future are equally accurate.  Across both uncued and character-cued responses (Fig.~\ref{fig:method}), retrodictions had numerically higher hit rates than predictions (Fig.~\ref{fig:result1}B).  However, these differences were only statistically reliable for character-cued responses (uncued responses: OR = 1.17, $Z = 0.35$, $p = 0.73$, CI: 0.47 to 2.92; character-cued responses: OR = 1.93, $Z = 2.15$, $p = 0.03$, CI: 1.06 to 3.52).  We observed a similar pattern of results for the precisions of participants' responses (Fig.~\ref{fig:result1}C).  Specifically, their responses tended to be numerically more precise for retrodictions versus predictions, but the differences were only statistically reliable for character-cued responses (uncued responses: $b = 0.03$, $t(20.9) = 1.09$, $p = 0.29$, CI: -0.03 to 0.10; character-cued responses: $b = 0.06$, $t(20.8) = 3.01$, $p = 0.007$, CI: 0.02 to 0.11).  We also consistently observed numerically higher convergence across participants for retrodictions versus predictions (Fig.~\ref{fig:result1}D), but neither of these differences were statistically reliable (uncued responses: $b = 0.03$, $t(17.9) = 0.75$, $p = 0.46$, CI: -0.05 to 0.11; character-cued responses: $b = 0.04$, $t(17.4) = 1.46$, $p = 0.16$, CI: -0.02 to 0.09).  Taken together, these results suggest that participants are generally better at making retrodictions than predictions.  We also verified that this was not solely a consequence of how participants' memory performance might have been affected by watching different segments (or making different responses to other segments) across conditions by comparing recall responses in the retrodiction-matched recall (\textit{re(R)}) and prediction-matched recall (\textit{re(P)}) conditions.  Recall performance was similar in both conditions (target event hit rate: OR = 1.12, $Z = 1.07$, $p = 0.29$, CI: 0.91 to 1.39; convergence: $b = 0.03$, $t(19.3) = 1.89$, $p = 0.07$, CI: 0.00 to 0.07).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{results2}
  \caption{\textbf{Retrodictions and predictions of temporally near and distant events.}  \textbf{A. Illustration of annotation approach.}  For each uncued retrodiction and prediction response, we calculated the number of (retrodicted or predicted) events as a function of temporal distance from the target segment, or \textit{lag}.  Onscreen (explicit) events are tagged using integer-valued lags, whereas offscreen (implicit) events are tagged using half-step lags ($\pm 0.5$, $\pm 1.5$, etc.).  \textbf{B. Number of events hit in participants' uncued retrodictions and predictions for each event type.}  Here we separated events we identified in participants' responses according to whether they occurred in the target segment (lags of $\pm 1$), during the interval between the target segment and the just-watched segment (lags of $\pm 0.5$), at longer temporal distances ($|\mathrm{lag}| > 1$), or were incorrect (unmatched with any past or future events in the narrative).  The counts displayed in the panel are averaged across just-watched segments.  \textbf{C. Number of events hit as a function of temporal distance.}  Here the (across-segment) mean numbers of events hit in participants' uncued retrodictions (orange) and predictions (blue) are displayed as a function of temporal distance to the just-watched segment (lag).  Error bars denote bootstrapped 95\% confidence intervals.  Colors denote temporal direction (orange: past; blue: future) and distance (darker shading: onscreen events from segments adjacent to the target segment; lighter shading: offscreen events).}
  \label{fig:result2}
\end{figure}

@@

Because characters tended to refer to past events more often than future events, the proportions of unreferenced events that were adjacent to referenced events should show a similar temporal asymmetry in favor of the past.  We tested this intuition by computing the proportions of unreferenced events in the stimulus that were temporally adjacent to past or future events referenced by the characters during a given segment.  Here we defined \textit{temporally adjacent} as any event within an absolute lag of one relative to a referenced onscreen event, or within an absolute lag of 0.5 to a referenced offscreen event.  We also defined \textit{remaining} events as unreferenced events that were not temporally adjacent to any referenced events.  As shown in Figure~\ref{fig:result4}B, we observed higher proportions of unreferenced past than future events that were temporally adjacent to referenced events.  Further, these reference-adjacent events had higher hit rates than remaining events after controlling for absolute lag (uncued retrodiction: OR = 7.15, $Z = 2.40$, $p = 0.02$, CI: 1.44 to 35.58; uncued prediction: OR = 3.11, $Z = 2.30$, $p = 0.02$, CI: 1.18 to 8.21; Fig.~\ref{fig:result4}E).  To estimate the contributions of reference adjacency on hit rates, we computed the difference in hit rates between unreferenced events (which comprised both reference-adjacent and remaining events) and remaining events, as a function of lag.  These differences exhibited a temporal asymmetry in favor of retrodiction.  This suggests that reference-adjacent events also contribute to participants' retrodiction advantage.  Remaining events did \textit{not} exhibit a reliable temporal asymmetry (OR = 0.75, $Z = 0.33$, $p = 0.74$, CI: 0.14 to 4.08; Fig.~\ref{fig:result4}D), suggesting that, after accounting for temporal adjacency, character's references to past and future events can explain participants' retrodiction advantage.

>>>>

\title{\Large The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events}
\author[1]{Xinming Xu}
\author[2]{Ziyan Zhu}
\author[3]{Xueyao Zheng}
\author[1, $\star$]{Jeremy R. Manning}
\affil[1]{Dartmouth College, Hanover, NH, USA}
\affil[2]{Peking University, Beijing, China}
\affil[3]{University of California at Davis, CA, USA}
\affil[$\star$]{Address correspondence to jeremy.r.manning@dartmouth.edu}

\newcommand{\stimDescription}{S1} % table
\newcommand{\segments}{S1} % figure
\newcommand{\events}{S2} %
\newcommand{\references}{S3}
\newcommand{\MethodsReplExp}{S4}
\newcommand{\targetAsymmetries}{S5}
\newcommand{\hitRates}{S6}
\newcommand{\characterRefs}{S7}
\newcommand{\refAdjacent}{S8}

@@

\begin{abstract} {\footnotesize{
  How much can we infer about the past and future, given our knowledge of the present?  Unlike temporally symmetric inferences about simple sequences, inferences about our own lives are asymmetric: we are better able to infer the past than the future, since we remember our past but not our future (i.e., the psychological arrow of time).  What happens when both the past and future are unobserved, as when we make inferences about \textit{other} people's lives?  We had participants in two experiments view segments of two character-driven television dramas.  They wrote out what would happen just before or after each just-watched segment.  Participants were better at inferring past (versus future) events.  This asymmetry was driven by participants’ reliance on characters’ conversational references in the narrative, which tended to favor the past.  We also carried out a meta analysis to estimate the prevalence of these asymmetries in hundreds of millions of dialogues from television shows, popular movies, novels, and written and spoken natural conversations. We found that, on average, references to the past are roughly 1.5--2 times more prevalent than references to the future.  Our work reveals a temporal asymmetry in how observations of other people’s behaviors can inform us about the past and future.

\textbf{Keywords: arrow of time, prediction, retrodiction, narrative, conversation}}}

@@

One key difference between simple artificial sequences and more naturalistic (real or narrative) sequences is that naturalistic sequences often incorporate other people.  Despite the past and future being equally unknown to \textit{the observer} prior to the current moment, other people, and realistic characters in narratives, have their own psychological arrows of time. Specifically, they have memories of their own pasts. Other people's asymmetric knowledge about their \textit{own} pasts and futures might affect their behaviors (e.g., conversations). In turn, this might provide time-asymmetric clues that favor the past~\citep[e.g., other people might talk more about their own pasts than their futures; ][]{DemiEtal18}. If observers leverage these clues from other people's asymmetric knowledge, then observers should also be better at inferring the past (versus the future) of other people's lives.  Alternatively, if inferences about other people's lives are more like inferences about artificial statistical sequences~\citep[e.g., perhaps solely relying on statistical regularities like event schemas, scripts, or situation models][]{RadvCope06, ZwaaRadv98, BoweEtal79, RangRitc12, BaldEtal18}, then the accuracy of inferences about the past and the future of others' lives should be approximately equal.

We designed a naturalistic paradigm for exposing participants to scenarios where the past and future were equally unobserved.  We asked our participants to watch a series of movie segments drawn from a character-driven dramatic television show.  Across the conditions and trials in the experiment, participants made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment.  We used manual annotations and sentence-level natural language processing models to characterize participants' responses.  To foreshadow our results, we found that participants were overall better at retrodicting the past than predicting the future. This appeared to be driven by two main factors.  First, characters more often referred to past events than future (e.g., planned) events, and this influenced participants' responses. Second, associations and dependencies between temporally adjacent events enabled participants to form estimates about nearby events (e.g., to a just-watched scene or a past or future event referenced in an observed conversation).  We also ran a pre-registered replication study to confirm that these findings generalized to another television show and group of participants.  Finally, we ran a meta analysis using natural language processing to estimate the prevelance of references to past and future events in hundreds of millions of dialogues drawn from television shows, popular movies, novels, and written and spoken natural conversations.  Taken together, our work reveals a temporal asymmetry in how observations of other humans’ behaviors inform us about the past versus the future.

\section*{Results}
Participants in our main experiment ($n = 36$) watched segments from two storylines, drawn from the CBS television show \textit{Why Women Kill}.  Each storyline comprised 11 segments (mean duration: 2.05~min; range: 0.97--3.87~min, Table~\stimDescription).  We asked participants to use free-form (typed) text responses to retrodict what had happened prior to a just-watched segment, predict what would happen next, or recall what they had just watched (Fig.~\ref{fig:method}, \textit{Task design}). We referred to the to-be-retrodicted, to-be-predicted, or to-be-recalled segment as the \textit{target segment} for each response. We systematically varied whether participants watched the segments in forward or reverse chronological order, and how many segments they had seen prior to making a response (see \textit{Methods}).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{methods}
  \caption{\textbf{Task overview.} Participants in our main experiment watched segments of two storylines from the television series \textit{Why Women Kill}.  They made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment. Across four counterbalanced sequences, we systematically varied whether participants watched the segments in forward or reverse chronological order, whether (or not) responses were cued using the main characters in the target segment, and which other segments participants had watched prior to making a response.  For each segment, we collected several retrodiction, prediction, and/or recall responses across different experimental conditions.  Experiment time is denoted along the
  vertical axis, storyline segments are indicated along the horizontal axis, and the colors denote experimental tasks (conditions).  For an analogous depiction of our replication experiment's design, see Fig.~\MethodsReplExp.}
  \label{fig:method}
\end{figure}

We asked participants in our main experiment to generate four types of responses after watching each video segment: uncued responses, character-cued responses, updated responses, and recalls (Fig.~\ref{fig:method}, \textit{Data overview}).  To generate \textit{uncued} responses, we asked participants to either retrodict (uncued retrodiction; \textit{u-R}) what happened shortly before or predict (uncued prediction; \textit{u-P}) what happened shortly after the just-watched segment.  To generate \textit{character-cued} responses, we asked participants to retrodict (character-cued retrodiction; \textit{c-R}) or predict (character-cued prediction; \textit{c-P}) what came before or after the just-watched segment, but we provided additional information to the participant about which character(s) would be present in the target (to-be-retrodicted or to-be-predicted) segment.  We hypothesized that character-cued responses should be more accurate than uncued responses, to the extent that participants incorporate the character information we provided to them into their retrodictions and predictions.  To generate updated responses, we asked participants to watch an additional segment that came just prior to or just after the target segment, and then to update their retrodiction (\textit{c-RP}) or prediction (\textit{c-PR}) about the target segment. Results on updated responses are not reported in this paper. Finally, we also asked participants to \textit{recall} what happened in the just-watched segment.  We labeled these responses according to which other segments participants had watched prior to the just-watched target.  Retrodiction-matched recall (\textit{re(R)}) responses were made during the retrodiction sequences (B1 and B2; Fig.~\ref{fig:method}), whereas prediction-matched recall (\textit{re(P)}) responses were made during the prediction sequences (A1 and A2; Fig.~\ref{fig:method}).  Whereas retrodiction and prediction responses reflect what participants \textit{estimate} they would remember after watching the (inferred) target segment, recall responses provide a benchmark for comparison by measuring what they \textit{actually} remember about the target segment.  Our replication experiment (Fig.~\MethodsReplExp) used a similar design, but did not have participants generate recall, re(R), or re(P) responses.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.80\textwidth]{results1}
  \caption{\textbf{Retrodiction, prediction, and recall performance by experimental condition.} \textbf{A.  Methods schematic.}  For each retrodiction, prediction, and recall response, we calculated the hit rate for events in the target segment, the response precision (see \textit{Methods}), and the response convergence across participants (see \textit{Methods}).  \textbf{B. Target event hit rate.} Mean proportions of target events that were contained in participants' responses, for each response type, averaged across target segments.  \textbf{C. Response precision.}  Mean precisions of participants' responses, for each response type, averaged across target segments.  The horizontal lines denote the mean pairwise semantic similarities (see \textit{Methods}) across recall responses (re(R): orange; re(P): blue).  \textbf{D. Response convergence.}  Mean (across-participant) convergence of participants' responses, for each response type, averaged across target segments.  All panels: error bars denote bootstrapped 95\% confidence intervals.  Asterisks indicate significance in the (generalized) linear mixed models: * denotes $p < 0.05$ and ** denotes $p < 0.01$.  See Figure~\targetAsymmetries for analogous results from our replication experiment.}
  \label{fig:result1}
\end{figure}

We used two general approaches to assess the quality of participants' responses (see \textit{Methods}, Figs.~\ref{fig:result1}A).  One approach entailed manually annotating events in the video and counting the number of matched events in participants' responses.  We identified a total of 117 unique events reflected across the 22 video segments (range: 3--9 per segment; see \textit{Methods}, Table \stimDescription).  We assigned one ``point'' to each of these video events.  We also identified 23 additional events in participants' responses that were either summaries of several events or that were partial matches to the manually identified video events.  We assigned 0.5 point to each of these additional events.  This point system enabled us to compute the numbers and proportions (\textit{hit rates}) of correctly retrodicted, predicted, and recalled events contained in each response.  Our second approach entailed using a natural language processing model~\citep{CerEtal18} to embed annotations and responses in a 512-dimensional feature space.  This approach was designed to capture conceptual overlap between responses that were not necessarily tied to specific events.  To quantify this conceptual overlap, we computed the similarities between the embeddings of different sets of responses.  Following \cite{HeusEtal21}, we defined the \textit{precision} of each participants' retrodictions or predictions about a target segment as the median cosine similarities between the embeddings of (a) the participant's retrodiction or prediction response for the target segment and (b) each \textit{other} participant's recalls of the same segment.  In other words, precision is designed to measure the extent to which retrodictions and predictions captured the conceptual content that (other) participants remembered.  We also developed a related measure, which we call \textit{convergence}, to characterize response similarities across participants.  In particular, we defined convergence as the mean cosine similarity between the embeddings of a participant's responses to a target segment and all other participants' responses (of the same type) to the same segment.  We analyzed the data using generalized linear mixed models, with participant and stimulus (e.g., target segment) identities as crossed random effects (see \textit{Methods}).

First we sought to validate a main effect of response type (i.e., uncued responses, character-cued responses, and recalls), irrespective of the temporal direction (retrodiction versus prediction).  Across these three types of responses, participants have access to increasing amounts of information about the target segment.  Therefore, across these response types, we hypothesized that participants' responses should become both more accurate and more convergent across individuals.  Consistent with this hypothesis, participants' character-cued retrodictions and predictions were associated with higher target event hit rates than uncued retrodictions and predictions (odds ratio (OR): 2.65, $Z = 4.24$, $p < 0.001$, 95\% confidence interval (CI): 1.69 to 4.16; Fig.~\ref{fig:result1}B).  These character-cued responses were also more precise ($b = 0.13$, $t(18.1) = 9.43$, $p < 0.001$, CI: 0.10 to 0.16; Fig.~\ref{fig:result1}C) and convergent across individuals ($b = 0.11$, $t(18.6) = 6.21$, $p < 0.001$, CI: 0.07 to 0.15; Fig.~\ref{fig:result1}D).   Relative to character-cued responses, participants' recalls showed higher target event hit rates (OR = 21.83, $Z = 10.61$, $p < 0.001$, CI: 12.35 to 38.59) and were more convergence across individuals ($b = 0.20$, $t(19.4) = 9.10$, $p < 0.001$, CI: 0.16 to 0.25). These results are consistent with the common-sense notion that access to more information about a target segment yields better performance (i.e., higher hit rates, precision, and convergence across individuals).  These findings also held for our replication experiment (Fig.~\targetAsymmetries; hit rates of character-cued vs. uncued responses: OR: XXX, $Z = XXX$, $p = XXX$, 95\% confidence interval (CI): XXX to XXX; precisions of character-cued vs. uncued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX; convergence of character-cued vs. uncued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX).

Next we carried out a series of analyses specifically aimed at characterizing temporal direction effects--- i.e, the relative quality of retrodictions versus predictions across different types of responses.  We hoped that these analyses might provide insights into our central question about whether inferences about the past and future are equally accurate.  Across both uncued and character-cued responses in our main experiment (Fig.~\ref{fig:method}), retrodictions had numerically higher hit rates than predictions (Fig.~\ref{fig:result1}B).  However, these differences were only statistically reliable for character-cued responses (uncued responses: OR = 1.17, $Z = 0.35$, $p = 0.73$, CI: 0.47 to 2.92; character-cued responses: OR = 1.93, $Z = 2.15$, $p = 0.03$, CI: 1.06 to 3.52).  We observed a similar pattern of results for the precisions of participants' responses (Fig.~\ref{fig:result1}C).  Specifically, their responses tended to be numerically more precise for retrodictions versus predictions, but the differences were only statistically reliable for character-cued responses (uncued responses: $b = 0.03$, $t(20.9) = 1.09$, $p = 0.29$, CI: -0.03 to 0.10; character-cued responses: $b = 0.06$, $t(20.8) = 3.01$, $p = 0.007$, CI: 0.02 to 0.11).  We also consistently observed numerically higher convergence across participants for retrodictions versus predictions (Fig.~\ref{fig:result1}D), but neither of these differences were statistically reliable (uncued responses: $b = 0.03$, $t(17.9) = 0.75$, $p = 0.46$, CI: -0.05 to 0.11; character-cued responses: $b = 0.04$, $t(17.4) = 1.46$, $p = 0.16$, CI: -0.02 to 0.09).  In our replication experiment (Fig.~\targetAsymmetries), participants were numerically better at making \textit{predictions} than retrodictions, but none of these differences were statistically reliable (hit rate for uncued responses: OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX; hit rate for character-cued responses: OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX; precision for uncued response: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX; precision for character-cued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX; convergence for uncued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX; convergence for character-cued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX).  Taken together, our results across our main and replication experiment suggest that whether participants are better at retrodicting versus predicting the immediate past or future may be somewhat stimulus specific.  We also verified that this was not solely a consequence of how participants' memory performance might have been affected by watching different segments (or making different responses to other segments) across conditions by comparing recall responses in the retrodiction-matched recall (\textit{re(R)}) and prediction-matched recall (\textit{re(P)}) conditions.  Recall performance in our main experiment was similar in both conditions (target event hit rate: OR = 1.12, $Z = 1.07$, $p = 0.29$, CI: 0.91 to 1.39; convergence: $b = 0.03$, $t(19.3) = 1.89$, $p = 0.07$, CI: 0.00 to 0.07).  (We did not collect recall responses in our replication experiment.)

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{results2}
  \caption{\textbf{Retrodictions and predictions of temporally near and distant events.}  \textbf{A. Illustration of annotation approach.}  For each uncued retrodiction and prediction response in our main experiment, we calculated the number of (retrodicted or predicted) events as a function of temporal distance from the target segment, or \textit{lag}.  Onscreen (explicit) events are tagged using integer-valued lags, whereas offscreen (implicit) events are tagged using half-step lags ($\pm 0.5$, $\pm 1.5$, etc.).  \textbf{B. Number of events hit in participants' uncued retrodictions and predictions for each event type.}  Here we separated events we identified in participants' responses according to whether they occurred in the target segment (lags of $\pm 1$), during the interval between the target segment and the just-watched segment (lags of $\pm 0.5$), at longer temporal distances ($|\mathrm{lag}| > 1$), or were incorrect (unmatched with any past or future events in the narrative).  The counts displayed in the panel are averaged across just-watched segments.  \textbf{C. Number of events hit as a function of temporal distance.}  Here the (across-segment) mean numbers of events hit in participants' uncued retrodictions (orange) and predictions (blue) are displayed as a function of temporal distance to the just-watched segment (lag).  Error bars denote bootstrapped 95\% confidence intervals.  Colors denote temporal direction (orange: past; blue: future) and distance (darker shading: onscreen events from segments adjacent to the target segment; lighter shading: offscreen events).  See Figure~\hitRates for an analogous presentation of results from our replication study.}
  \label{fig:result2}
\end{figure}

@@

Because characters tended to refer to past events more often than future
events, the proportions of unreferenced events that were adjacent to referenced
events should show a similar temporal asymmetry in favor of the past. We tested
this intuition by computing the proportions of unreferenced events in the
stimulus that were temporally adjacent to past or future events referenced by
the characters during a given segment. Here we defined \textit{temporally
adjacent} as any event within an absolute lag of one relative to a referenced
onscreen event, or within an absolute lag of 0.5 to a referenced offscreen
event. We also defined \textit{remaining} events as unreferenced events that
were not temporally adjacent to any referenced events. As shown in
Figure~\ref{fig:result4}B, in our main experiment we observed higher
proportions of unreferenced past than future events that were temporally
adjacent to referenced events. Further, these reference-adjacent events had
higher hit rates than remaining events after controlling for absolute lag
(uncued retrodiction: OR = 7.15, $Z = 2.40$, $p = 0.02$, CI: 1.44 to 35.58;
uncued prediction: OR = 3.11, $Z = 2.30$, $p = 0.02$, CI: 1.18 to 8.21;
Fig.~\ref{fig:result4}E). These findings also held in our replication
experiment (uncued retrodiction: OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to
XXX; uncued prediction: OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX;
Fig.~\refAdjacent). To estimate the contributions of reference adjacency on hit
rates, we computed the difference in hit rates between unreferenced events
(which comprised both reference-adjacent and remaining events) and remaining
events, as a function of lag. These differences exhibited a temporal asymmetry
in favor of retrodiction. This suggests that reference-adjacent events also
contribute to participants' retrodiction advantage. Remaining events did
\textit{not} exhibit a reliable temporal asymmetry (main experiment: OR = 0.75,
$Z = 0.33$, $p = 0.74$, CI: 0.14 to 4.08, Fig.~\ref{fig:result4}D; replication
experiment: OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX, Fig.~\refAdjacent
D), suggesting that, after accounting for temporal adjacency, character's
references to past and future events can explain participants' retrodiction
advantage.

\textbf{JRM STOPPED HERE...}
